---
title: "Template"
author: "Jakob, Ben, and David"
date: "`r Sys.Date()`"
output: pdf_document
bibliography: references.bib
csl: apa.csl

---

# Set-up your environment

```{r package_install, include=FALSE}
# install.packages("tidyverse")
# install.packages("yaml")
# install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(tidyr)
library(writexl)
library(readxl)
```
```{r}
gdp_data           <- read_csv("data/raw_gdp.csv")
unemployment_sex   <- read_csv("data/raw_unemployment_sex.csv")
unemployment_youth <- read_csv("data/raw_unemployment_youth.csv")
```
```{r}
# Filter all three datasets down to just the Netherlands
gdp_clean   <- gdp_data %>% filter(geo == "Netherlands")
unemp_clean <- unemployment_sex %>% filter(geo == "Netherlands")
youth_clean <- unemployment_youth %>% filter(geo == "Netherlands")
```
```{r}
# Plot both general and youth lines together using your perfect 11-row sheets
ggplot() +
  geom_line(data = unemp_clean, aes(x = TIME_PERIOD, y = OBS_VALUE, color = "Total Unemployment"), size = 1) +
  geom_line(data = youth_clean, aes(x = TIME_PERIOD, y = OBS_VALUE, color = "Youth Unemployment"), size = 1) +
  theme_minimal() +
  scale_color_manual(values = c("Total Unemployment" = "darkred", "Youth Unemployment" = "orange")) +
  labs(
    title = "Unemployment Comparison in the Netherlands (2015-2025)",
    x = "Year",
    y = "Unemployment Rate (%)",
    color = "Demographic Group"
  )
```
```{r}
# 1. Combine the datasets safely
combined_data <- left_join(gdp_clean, unemp_clean, by = "TIME_PERIOD")

# 2. Draw the Okun's Law chart with explicit chronological labels
ggplot(data = combined_data, aes(x = OBS_VALUE.x, y = OBS_VALUE.y)) +
  geom_point(size = 4, color = "darkblue") +
  
  # MAGIC LAYER: This punches the timeline back onto the canvas
  geom_text(aes(label = TIME_PERIOD), vjust = -1.2, size = 3.5, fontface = "bold", color = "black") +
  
  geom_smooth(method = "lm", color = "red", linetype = "dashed", se = FALSE) +
  theme_minimal() +
  labs(
    title = "Okun's Law: GDP Growth vs. Unemployment Timeline",
    subtitle = "Text labels reveal chronological policy distortions and labor market lags",
    x = "Annual GDP Growth Rate (%)",
    y = "Total Unemployment Rate (%)"
  )
```
```{r}
# Event Analysis: Total Unemployment with a 2020 Shock Policy Line
ggplot(data = unemp_clean, aes(x = TIME_PERIOD, y = OBS_VALUE)) +
  geom_line(color = "darkred", size = 1) +
  geom_point(color = "darkred", size = 2.5) +
  # This adds the vertical dashed line required by the rubric
  geom_vline(xintercept = 2020, linetype = "dashed", color = "black", size = 1) +
  # This places the text label directly next to the line
  annotate("text", x = 2020.2, y = 7, label = "COVID-19 Shock (2020)", hjust = 0, fontface = "bold") +
  theme_minimal() +
  labs(
    title = "Event Analysis: Impact of the 2020 Shock on Dutch Labor",
    x = "Year",
    y = "Unemployment Rate (%)"
  )
```
```{r,message=FALSE, warning=FALSE}
library(maps)

# 1. Pull European map coordinates from R's built-in database
europe_map <- map_data("world") %>%
  filter(region %in% c("Netherlands", "Germany", "Belgium", "France", "Italy", "Spain", "Austria", "Denmark"))

# 2. Extract 2020 data directly without unnecessary sub-filters
europe_unemp_2020 <- unemployment_sex %>%
  filter(TIME_PERIOD == 2020) %>%
  rename(region = geo) 

# 3. Merge the coordinates and the economic data together
map_data_joined <- left_join(europe_map, europe_unemp_2020, by = "region")

# 4. Draw the spatial map with the color gradient filled in
ggplot(map_data_joined, aes(x = long, y = lat, group = group, fill = OBS_VALUE)) +
  geom_polygon(color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "gray80", name = "Unemployment (%)") +
  coord_fixed(1.3) +
  theme_minimal() +
  labs(
    title = "Spatial Variation: Unemployment Rates Across Western Europe (2020)",
    subtitle = "Visualizing regional labor market vulnerability during the pandemic shock",
    x = "Longitude",
    y = "Latitude"
  )
```

# Title Page
A new sentence
Include your names

Include the tutorial group number

Include your tutorial lecturer's name

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant? [@smith2020]

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
#dataset <- midwest

library(readr)
#Gezondheid_leefstijl_zorggebruik_en_aanbod_doodsoorzaken_kerncijfers_03062026_132800 <- #read_delim("~/Downloads/Gezondheid__leefstijl__zorggebruik_en__aanbod__doodsoorzaken__kerncijfers_03062026_132800.csv", 
#    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
#    skip = 4)
#View(Gezondheid_leefstijl_zorggebruik_en_aanbod_doodsoorzaken_kerncijfers_03062026_132800)

#write_csv(Gezondheid_leefstijl_zorggebruik_en_aanbod_doodsoorzaken_kerncijfers_03062026_132800, "data/gezondheid.csv")

a_fun_name <- read_delim("data/gezondheid.csv")

a_fun_name$...1 <- NULL


write_csv(a_fun_name, "data/gezondheid_changed.csv")
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
